<<<<<<< HEAD
---
title: "Data Science Methods - Assignment 2"
author:
- "M. Alberti, s.n. 2020162"
- "N. Ceschin, s.n. 344510"

date: March 16, 2020
output: pdf_document

---

## Question 1

First we upload all relevant libraries:

```{r results='hide', message=FALSE, warning=FALSE}
rm(list=ls())

# load libraries
library(leaps)
library(glmnet) # for lasso and ridge
library(Matrix)
library(pROC)
```

Data preparation

```{r results='hide', message=FALSE, warning=FALSE}
setwd("C:/Users/nicol/Documents/Tilburg/Courses/Data Science Methods/DATA-SCIENCE-ASSIGNMENTS/Assignment 2")
data<- read.csv("R_class.csv", sep=",", dec=".", header=TRUE)

ca <- grep("ca", names(data), value=T)  #vector of names of columns containing ca
drops <- names(data) %in% c(ca)
data <- data[!drops]  

# drop vars not used
assets <- grep("assets", names(data), value=T)
stocks <- grep("stocks", names(data), value=T)
narrowm <- grep("narrowm", names(data), value=T)
money <- grep("money", names(data), value=T)
ltrate <- grep("ltrate", names(data), value=T)
stir <- grep("stir", names(data), value=T)
loans <- grep("loans", names(data), value=T)
debt <- grep("debt", names(data), value=T)
er <- grep("er", names(data), value=T)
cpi <- grep("cpi", names(data), value=T)
gap <- grep("gap", names(data), value=T)
glo <- grep("a_", names(data), value=T)
gdp <- grep("gdp", names(data), value=T)
i <- grep("i_", names(data), value=T)
c <- grep("c_", names(data), value=T)
ri <- grep("ri", names(data), value=T)
rc <- grep("rc", names(data), value=T)

drops <- names(data) %in% c("year", "ccode", stocks, money, stir,assets,i,ri,glo) 
saves <- names(data) %in% c(glo)
full <- data[!drops] # drops those variables which have true indication in "drops"
full <- cbind(data[glo], full)

# FULL SET: omit observations with missing values
full_om <- na.omit(full)
```

Ridge regression

```{r warning=FALSE}
#drop alternative independent variables
misc.list <- c("b1","b3","rec1","rec2","rec3")
location <- names(full_om) %in% c(misc.list) 
name.indep <- names(full_om[!location])
indep <- full_om[name.indep]

#division of the sample
set.seed(4)
indexes = sample(1:nrow(full_om), size=0.632*nrow(full), replace=F)
test = full_om[-indexes,]
train = full_om[indexes,]

#Run ridge
train.mat = model.matrix(b2~., data=train) 
test.mat = model.matrix(b2~., data=test)
grid = 10 ^ seq(2, -3, length=100)
mod.ridge = cv.glmnet(train.mat, train[, "b2"], alpha=0, lambda=grid, family="binomial", thresh=1e-12) 
lambda.best = mod.ridge$lambda.min 
ridge.probs = predict(mod.ridge, newx=test.mat, s=lambda.best, type="response")
true <- test[,"b2"]
r<-roc(true, ridge.probs, ci=T) # ROC analysis
###aucs[1,j] <- as.numeric(r$auc)		
	#ci95_lo[1,j] <- as.numeric(ci.auc(r,conf.level=ci[2]))[1]
	#ci95_up[1,j] <- as.numeric(ci.auc(r,conf.level=ci[2]))[3]


```
=======
---
title: "Data Science Methods - Assignment 2"
author:
- "M. Alberti, s.n. 2020162"
- "N. Ceschin, s.n. 344510"

date: March 16, 2020
output: pdf_document

---

## Question 1

First we upload all relevant libraries:

```{r results='hide', message=FALSE, warning=FALSE}
rm(list=ls())

# load libraries
library(leaps)
library(glmnet) # for lasso and ridge
library(Matrix)
```

Data preparation

```{r results='hide', message=FALSE, warning=FALSE}
setwd("C:/Users/nicol/Documents/Tilburg/Courses/Data Science Methods/DATA-SCIENCE-ASSIGNMENTS/Assignment 2")
data<- read.csv("R_class.csv", sep=",", dec=".", header=TRUE)

ca <- grep("ca", names(data), value=T)  #vector of names of columns containing ca
drops <- names(data) %in% c(ca)
data <- data[!drops]  

# drop vars not used
assets <- grep("assets", names(data), value=T)
stocks <- grep("stocks", names(data), value=T)
narrowm <- grep("narrowm", names(data), value=T)
money <- grep("money", names(data), value=T)
ltrate <- grep("ltrate", names(data), value=T)
stir <- grep("stir", names(data), value=T)
loans <- grep("loans", names(data), value=T)
debt <- grep("debt", names(data), value=T)
er <- grep("er", names(data), value=T)
cpi <- grep("cpi", names(data), value=T)
gap <- grep("gap", names(data), value=T)
glo <- grep("a_", names(data), value=T)
gdp <- grep("gdp", names(data), value=T)
i <- grep("i_", names(data), value=T)
c <- grep("c_", names(data), value=T)
ri <- grep("ri", names(data), value=T)
rc <- grep("rc", names(data), value=T)

drops <- names(data) %in% c("year", "ccode", stocks, money, stir,assets,i,ri,glo) 
saves <- names(data) %in% c(glo)
full <- data[!drops] # drops those variables which have true indication in "drops"
full <- cbind(data[glo], full)

# FULL SET: omit observations with missing values
full_om <- na.omit(full)
```




>>>>>>> 37e282b41b2b3b6b1e0ea1404fc551e6ee196297
